---
title: "Cleaning and Analyzing the Penguins Dataset"
author: "Korey W"
format: html
editor: visual
---

## **Introduction**
This document processes the `penguins_raw_dirty.csv` dataset.  
It first **runs an external R script** to clean the data, then performs basic **data analysis and visualization**.

---

## **Step 1: Run the Cleaning Script**
```{r}
source("clean_penguins.R")
```
This **runs `clean_penguins.R`**, which:  
- Cleans the dataset  
- Saves it as `penguins_cleaned.csv` in `Processed_data/`  

---

## **Step 2: Load the Cleaned Data**
```{r}
# Load necessary libraries
library(tidyverse)
library(here)

# Load cleaned data
cleaned_data <- read.csv(here("Data", "Processed_data", "penguins_cleaned.csv"))

# Preview the first few rows
head(cleaned_data)

# Check actual column names
colnames(cleaned_data)

# Rename the columns correctly based on colnames output
cleaned_data <- cleaned_data %>%
  rename(
    Flipper_Length_mm = Flipper_Length_mm_,  
    Body_Mass_g = Body_Mass_g_ 
)

# Standardize species names (fix typos and inconsistencies)
cleaned_data <- cleaned_data %>%
  mutate(Species = case_when(
    str_detect(Species, "Adelie|AdeKie|PengTin|Pengufn|PeOguin|MPenguin") ~ "Adelie",
    str_detect(Species, "Gentoo|Ventoo") ~ "Gentoo",
    str_detect(Species, "Chinstrap") ~ "Chinstrap",
    TRUE ~ Species  # Keep unchanged if already correct
  ))


```
This **loads the cleaned dataset** and **renames columns**.

---

## **Step 3: Data Summary**
```{r}
# Summary of all numeric columns
summary(cleaned_data)

# Check unique values for categorical variables
unique(cleaned_data$Species)
unique(cleaned_data$Island)
unique(cleaned_data$Sex)
```
This **shows basic statistics and checks for typos in categories**.

---

## **Step 4: Data Visualization**

### **Species Distribution**
```{r}
cleaned_data %>%
  count(Species) %>%
  ggplot(aes(x = Species, y = n, fill = Species)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Count of Penguins by Species", y = "Count", x = "Species")
```
This **creates a bar chart** showing how many of each species are in the dataset.

---

### **Body Mass vs. Flipper Length**
```{r}
ggplot(cleaned_data, aes(x = Flipper_Length_mm, y = Body_Mass_g, color = Species)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Penguin Body Mass vs. Flipper Length",
       x = "Flipper Length (mm)", y = "Body Mass (g)")
```
This **creates a scatter plot** comparing body mass and flipper length.

---

## **Conclusion**
This document:  
**Cleans the dataset** using `clean_penguins.R`  
**Loads the cleaned dataset**  
**Performs basic data summaries**  
**Visualizes key relationships**  

**Now the dataset is ready for further analysis!**

---

## **How to Run This**
Once you’ve saved this file as `clean_penguins_sourced.qmd`, run **this single command in Terminal**:
```sh
quarto render clean_penguins_sourced.qmd
```
This **runs everything**—you don’t need to switch between files anymore.  
The final output will be an **HTML report** in the same folder.
