---
title: "Cleaning the Penguin Data - an example"
author: "Marguerite Butler"
date: "2023-03-20"
output: html_document
---


# Processing script (data cleaning) with code pulled in from script

This quarto file documents the data cleaning of the raw data on penquins `penguins_raw_dirty.csv` in the raw data folder `../../Data/Raw_data/`. This script will produce a cleaned data file stored in the directory `../../Data/Processed_data`. The clean data file is named `processeddata.rds` and also saved in `.csv` formats.

To run this code, run `quarto preview processingfile_v2.qmd` from the Processing code directory `project_root/Code/Processing_code`.  Alternatively, you can prepend the relative path to the file from your current working directory. To generate the `.html` output, change  `quarto preview` to `quarto render`. 

# Setup

This code is needed pull the code chunks contained in `processingcode.R` into active memory:

```{r, include=TRUE, cache=FALSE}
knitr::read_chunk('processingcode.R')
```

Load needed packages (from the `packages` code chunk):

```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading

```{r, loaddata}
```

# Explore Data

### There are several ways of looking at the data

```{r, exploredata}
```

We also use short variable names (temporarily) in this script for convenience. We save the cleaned data with the original names, but save the short names at the end to reuse them later. 

# Cleaning

Inspecting the data, we find some problems that need addressing. 

###  Species names

First, we know that this is a dataset for three species of penguin, but we notice that there are 9 unique species.

```{r, cleandata1}
```

Fix all of the errors. 

Also, letʻs shorten Species just keeping the three common names "Adelie", "Gentoo", and "Chinstrap" and delete the rest of the Species character string. 

```{r, fixtypos}
```

### Continuous data

There is an entry for `Culmen Length (mm)` which says "missing" instead of a number or NA. 
Should we delete this record (and all of the variables)?
This "missing" entry also turned all culmen length entries into characters instead of numeric.
That conversion to character also means that our summary function isn't very meaningful.

So let's fix that first.

```{r, cleandata2}
```

Notice anything funny? 

Now we see that there are three penguins with really really long culmens (300+mm) that could be typos. If we don't know, we might need to remove these penguins. But let's suppose that we somehow know that this is because of a misplaced decimal point (for example if we could verify with field records), and letʻs fix this:

```{r, cleandata3.1}
```

Notice that NAʻs will match the condition `cl>300`, NAs will match ANY condition. R does this to be conservative because it is unknown whether it matches or does not match. We donʻt want NAs, so letʻs exclude them with 
!is.na()  where the ! is "not" or the opposite of is.na:

```{r, cleandata3.2}
```


Look better?

### Now let's look at body mass.

There are penguins with body mass of <100g when the others are over 3000. 
Perhaps these are new chicks? But they are supposed to be adults. Letʻs remove them.

```{r, cleandata4.1}
```


Mass is the main size variable, so we will remove the individuals with missing masses in order to  be able to analyze the data. 

Note: Some analysis methods can deal with missing values, so it's not always necessary to drop observations with missing data in the non-size variables. Here we will drop observations with missinging mass but leave in NAs in other variables. 

```{r, cleandata4.2}
```

These data now look cleaned of impossible values. 

### Factors

We also want to have Species, Sex, and Island coded as a categorical/factor variable:

```{r, cleandata5}
```

# Bivariate Plots

We look through all of the data, variable by variable, or by pairs of variables to check on the distributions of the data and for any potential errors in the data. 

We make bivariate plots for each continous variable to ensure there are no further errors. It is a good practice to check on the distribution of the data as well. 

```{r, bivariateplots}
```

# Histograms by Species

We make histograms or densities of mass by each species, to check for any potential category errors, extra categories, etc.  


```{r, histograms}
```


# Finalize the cleaned dataset. 

Drop any variables (columns) that we wonʻt analyze. We also make a new data dictionary to save with the cleaned data. 

```{r, finalizedata}
```
# Save data

All done, data is clean now. 
Finally, we save the clean data as an RDS file, as well as a copy as .csv for archival puproses.

```{r, savedata}
```




